Tropical mathematics for the Lambda-calculus 1

### 2023-07-28 CSL24/section2.tex
Commit serale

The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma \vdash M:A
		\justifies
		\Gamma, x:_{0}B \vdash M:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
		\justifies
		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n} A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
		\endprooftree
		&
		\prooftree
		\Gamma \vdash M: !_nA\multimap B
		\quad
		\Delta\vdash N: A
		\justifies
		\Gamma +n\Delta\vdash MN: B
		\endprooftree
	\end{array}
	\]
where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated at most $n$ times in the reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
%E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}. yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

%\begin{remark}\label{rmk:ModelsOfBSTLC}
%Similarly to It is known {\color{red}(([reference??] e dire meglio)} 
The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
Now, one can see that the exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials functors $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a $\N$-graded linear exponential comonad on (the SMC) $\LREL$, %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
i.e.\ $(\LREL,(!_n)_{n\in\N})$ is a model of $\BSTLC$. 
In particular, bounded arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and so, 
if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
%\end{remark}
%Bounded types are interesting because of the following proposition:
%\begin{proposition}
%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
%\end{proposition}
%\begin{proof}
%It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
%\end{proof}

@@

The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%This is crucial for the definition of the Taylor expansion.
We follow the tradition of quotienting also for the idempotency of $+$.
%Sums are, then, just \emph{finite} sets of terms.
The types are $A::= *\mid A\to A$, the typing rules: %in Figure~\ref{fig:rules}
	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma, x: A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: A\to B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma\vdash \mathbb T: A
		\justifies
		\Gamma \vdash M\mathbb T: B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma \vdash N: A
		\justifies
		\Gamma \vdash \Diff{M}{N}: A\to B
		\endprooftree
		&
		\prooftree
		\Gamma\vdash M_1: A
		\,\cdots\,
		\Gamma \vdash M_n:A
		\justifies
		\Gamma \vdash M_1+\cdots +M_n : A
		\using (n\geq 2)
		\endprooftree
	\end{array}
	\]
where a context $\Gamma$ is a list of typed variable declarations.
The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during the reduction to normal form.
%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
\begin{comment}
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}*}\to ({\color{green}* \to * \to *}) \to *$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
\end{comment}
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).

@@

%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}. 
The $\STDLC$ can be interpreted in Cartesian closed differential $\lambda$-categories (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019}.
In them, homsets are equipped with a structure of commutative monoid %(i.e.\ we can add morphisms and there is a $0$ morphism)
%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
and with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
Now, applying \cite[Theorem 6.1]{lemay2020} one can check that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ becomes a CC$\partial\lambda$C, i.e.\ a model of the $\STDLC$, when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).

As already mentioned, by induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
one can prove that:

\begin{proposition}\label{prop:descrete}
 For $\BSTLC$ and $\STDLC$ (but not $\STLC_\oplus$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
\end{proposition}

Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}

>>>>

One can see that the comonad $!$ of $\LREL$ can be ``decomposed'' {\color{red}REF} into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), defined on objects by taking multisets of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$. %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.

As such, $(\LREL,(!_n)_{n\in\N})$ is a model of graded calculi ensuring bounded duplications via graded typing systems, for instance it is a model  \cite{Katsumata2018} of the language $\BSTLC$, a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}, defined as follows: 
the terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$, the contexts of the typing judgements are sets of declarations of the form $x :_{n}A$, with $n\in \mathbb N$, and the typing rules are: %given in Fig.~\ref{fig:rules}
	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma \vdash M:A
		\justifies
		\Gamma, x:_{0}B \vdash M:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n}B, y:_{m} B\vdash M:A
		\justifies
		\Gamma, x:_{n+m}B\vdash M\{x/y\}:A
		\endprooftree 
		&
		\prooftree
		\Gamma, x:_{n} A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: !_{n}A\multimap B
		\endprooftree
		&
		\prooftree
		\Gamma \vdash M: !_nA\multimap B
		\quad
		\Delta\vdash N: A
		\justifies
		\Gamma +n\Delta\vdash MN: B
		\endprooftree
	\end{array}
	\]
where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ is duplicated at most $n$ times in the reduction of $\vdash (\lambda x.M)N :B$ to the normal form.
%E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}. yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

%\begin{remark}\label{rmk:ModelsOfBSTLC}
%Similarly to It is known {\color{red}(([reference??] e dire meglio)}

Remark that since arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$, 
if $\model *$ is finite, then $\model{A}$ is a finite set for any type $A$ of $\BSTLC$.
%\end{remark}
%Bounded types are interesting because of the following proposition:
%\begin{proposition}
%For all bounded types $A,B$, the morphisms from $\model A$ to $\model B$ (in all parametric relational models) correspond to polynomials.
%\end{proposition}
%\begin{proof}
%It suffices to check that $\model A$ is finite for all bounded types $A$. Indeed this implies that a morphism $t:\model A\to \model B$ is a finite matrix $t: \model A \times \model B \to \Lawv$.Hence, its corresponding map $\widehat t:\Lawv^{\model A} \to \Lawv^{\model B}$ is a polynomial.
%\end{proof}

@@

A Cartesian closed differential $\lambda$-category (CC$\partial\lambda$C)\cite{Manzo2010,Blute2009, Blute2019} is a CCC enriched over commutative monoids %(i.e.\ we can add morphisms and there is a $0$ morphism)
%, it is Cartesian closed%(with the closed structure compatible with the additive structure)
and equipped with a certain \emph{differential operator} $D$, generalising the usual notion of differential, see e.g.\ \cite{BluteEhrhTass10}.
%An example is the CC$\partial\lambda$C of convenient vector spaces with smooth maps, where $D$ is the ``real'' differential $Df:\mathbb{V}\times\mathbb{V}\rightarrow \mathbb{W}$, $Df(x,u):=\dfrac{d}{dt}{\!\Big|_{t=0}} f(x+tu)$, of smooth maps $f:\mathbb{V}\rightarrow \mathbb{W}$.
%More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
Now, applying \cite[Theorem 6.1]{lemay2020} one can check that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ becomes a CC$\partial\lambda$C when equipped with the \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).

As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}, a language ensuring exact control of duplications via a notion of linear substitution.
Its syntax (e.g.\ \cite[Section 3]{Manzo2010}) is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
quotiented by $\alpha$-equivalence, by equations that make $+,0$ form a commutative monoid on the set of sums, %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
by linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$) and by irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%This is crucial for the definition of the Taylor expansion.
We follow the tradition of quotienting also for the idempotency of $+$.
%Sums are, then, just \emph{finite} sets of terms.
The types are $A::= *\mid A\to A$, the typing rules: %in Figure~\ref{fig:rules}
	\[ \scriptsize \arraycolsep=5pt\def\arraystretch{2.8}
	\begin{array}{cccc}
		\prooftree
		\Gamma, x: A\vdash M: B
		\justifies
		\Gamma\vdash \lambda x.M: A\to B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma\vdash \mathbb T: A
		\justifies
		\Gamma \vdash M\mathbb T: B
		\endprooftree 
		&
		\prooftree
		\Gamma \vdash M: A\to B
		\quad
		\Gamma \vdash N: A
		\justifies
		\Gamma \vdash \Diff{M}{N}: A\to B
		\endprooftree
		&
		\prooftree
		\Gamma\vdash M_1: A
		\,\cdots\,
		\Gamma \vdash M_n:A
		\justifies
		\Gamma \vdash M_1+\cdots +M_n : A
		\using (n\geq 2)
		\endprooftree
	\end{array}
	\]
where a context $\Gamma$ is a list of typed variable declarations.
The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.
The main feature of this language is that $\Der^n[\lambda x.M,N^n]0$ has a non-zero normal form iff $x$ is duplicated exactly $n$ times during the reduction to normal form.
%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
\begin{comment}
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}*}\to ({\color{green}* \to * \to *}) \to *$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).
\end{comment}
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).

@@

%The models of $\STDLC$ are the cartesian \emph{closed} differential categories ($CC\partial C$), which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.

### 2023-07-28 CSL24/section2bis.tex
Commit serale

A \emph{tropical polynomial} is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$, where $i_{j}\in\N$ and $c_{i_{j}}\in\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
%For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=k\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

A \emph{tropical Laurent series} of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\{nx+\matr f_{n}\}$, with $\matr  f_{n}\in\Lawv$.
In other words, it is a ``limit'' of tropical polynomials of higher and higher degree.
E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
is the ``limit'' of the polynomials $\varphi_{n}$.
Since $\inf$s are not in general $\min$s, the behaviour of tLS may be less predictable than that of tropical polynomials~\cite{Porzio2021}. %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.

@@

Now, for a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.

>>>>

\begin{definition}
A \emph{tropical power series} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$ and $I\subseteq \N$.
If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
\end{definition}

A tropical polynomial is always a piece-wise linear function since it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes), or equivalently, where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
%For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
%Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=k\min\{x,x_{0}\}+ g(x)$.
Unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

Since $\inf$s are not in general $\min$s, a tPs is a ``limit'' of tropical polynomials of higher and higher degree, and its behaviour may be less predictable than that of tropical polynomials~\cite{Porzio2021}. %For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points of their domain.
E.g., the tLs $\varphi(x):=\inf_{n\in\N}\set{nx+2^{-n}}$ (see Fig.~\ref{fig:plot1}) %that we will take as our running example, 
is the ``limit'' of the polynomials $\varphi_{n}$.

@@

Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.

### 2023-07-28 CSL24/section3A.tex
Commit serale

The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once}. %(and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
Now, $\LREL$ is a SMCC (\cite[Section III.A]{Manzo2013}), %(actually, even a model of IMLL), 
whose monoidal product $\otimes$ acts on objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
 %{(\color{red} Ref?)} 
 and its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.

>>>>

$\LREL$ is a Symmetric Monoidal Closed Category (SMCC) \cite[Section III.A]{Manzo2013}, %(actually, even a model of IMLL), 
whose monoidal product $\otimes$ acts on objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
 %{(\color{red} Ref?)} 
 and its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.
As such, it is able to interpret the \emph{linear simply typed $\lam$-calculus}, the restriction of the ordinary $\lam$-calculus in which programs ,% following the linear logic mantra, 
can only use their arguments \emph{exactly once}. %(and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.

### 2023-07-28 CSL24/section3B.tex
Commit serale

\begin{remark}
One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
This is also the case for the differential $\lambda$-calculus that we will introduce in a moment, so we state and discuss this property at that point (\autoref{prop:descrete}).
%It is instructive what its CCC-structure looks like in our tropical world.
\end{remark}

>>>>

By induction on $M$ %(using that composition and $D$ easily preserve ``booleaness'' and projections and evaluation of  $\LREL$ are boolean)
one can prove that:

\begin{proposition}\label{prop:descrete}
 For $\BSTLC$ and $\STDLC$ (and thus also for its fragment $\STLC$), we have that $\model{\Gamma\vdash M:A}\in\Lawv^{!\model{\Gamma}\times\model{A}}$ is a \emph{boolean} matrix, i.e.\ actually $\model{\Gamma\vdash M:A}\in\{0,\infty\}^{!\model{\Gamma}\times\model{A}}$.
\end{proposition}

This can be seen as the fact that such calculi are relatively trivial, from this point of view.
Even if the interpretation of those languages is trivia, we shall still focus on them in order to set the basis of the further studies and in order to see some already interesting properties of their interpretation.{\color{red}SPIEGARE MEGLIO}
%It is instructive what its CCC-structure looks like in our tropical world.
